### 2023-11-13 lmcs/priority-gv.tex
Update lmcs.cls

We illustrate this with two programs in PGV, \cref{ex:cycle,ex:deadlock}. Each program contains two processes---the main process, and the child process created by $\tm{\spawn}$---which communicate using \emph{two} channels. The child process receives a unit over the channel $\tm{x}/\tm{x'}$, and then sends a unit over the channel $\tm{y}/\tm{y'}$. The main process does one of two things:
\begin{enumerate}[label= (\alph*) ]
\item in \cref{ex:cycle}, it sends a unit over the channel $\tm{x}/\tm{x'}$, and then waits to receive a unit over the channel $\tm{y}/\tm{y'}$;
\item in \cref{ex:deadlock}, it does these in the opposite order, which results in a deadlock.
\end{enumerate}
PGV is more expressive than GV: \cref{ex:cycle} is typeable and guaranteed to be deadlock-free in PGV, but is not typeable in GV~\cite{wadler14} and not guaranteed deadlock-free in GV's predecessor~\cite{gayvasconcelos10}. We believe PGV is a non-conservative \emph{extension} of GV, as CP can be embedded in a Kobayashi-style system~\cite{dardhaperez15extended}.

\begin{figure}
\begin{minipage}{0.5\linewidth}
\begin{example}[Cyclic Structure]%
  \label{ex:cycle}
  \[\tm{%
    \begin{array}{l}
      \letpair{x}{x'}{\new\;\unit}{}
      \\
      \letpair{y}{y'}{\new\;\unit}{}
      \\
      \spawn\left(\lambda\unit.%
      \begin{array}{l}
        {\letpair{\unit}{x'}{\recv\;{x'}}{}}
        \\
        {\letbind{y}{\send\;{\pair{\unit}{y}}}{}}
        \\
        {\andthen{\wait\;{x'}}{\close\;{y}}}
      \end{array}%
      \right);
      \\
      \underline{\letbind{x}{\send\;{\pair{\unit}{x}}}{}}
      \\
      \underline{\letpair{\unit}{y'}{\recv\;{y'}}{}}
      \\
      {\andthen{\close\;{x}}{\wait\;{y'}}}
    \end{array}
  }\]
\end{example}
\end{minipage}%
\begin{minipage}{0.5\linewidth}
\begin{example}[Deadlock]%
  \label{ex:deadlock}
  \[\tm{%
    \begin{array}{l}
      \letpair{x}{x'}{\new\;\unit}{}
      \\
      \letpair{y}{y'}{\new\;\unit}{}
      \\
      \spawn\left(\lambda\unit.%
      \begin{array}{l}
        {\letpair{\unit}{x'}{\recv\;{x'}}{}}
        \\
        {\letbind{y}{\send\;{\pair{\unit}{y}}}{}}
        \\
        {\andthen{\wait\;{x'}}{\close\;{y}}}
      \end{array}%
      \right);
      \\
      \underline{\letpair{\unit}{y'}{\recv\;{y'}}{}}
      \\
      \underline{\letbind{x}{\send\;{\pair{\unit}{x}}}{}}
      \\
      {\andthen{\close\;{x}}{\wait\;{y'}}}
    \end{array}
   }\]
\end{example}
\end{minipage}
\end{figure}

@@

We must show that substitution preserves priority constraints. For this, we prove~\cref{lem:pgv-value-done}, which shows that values have finished all their communication, and that any priorities in the type of the value come from the typing environment.
%The proofs are by structural induction and can be found in \cref{prf:lem-pgv-value-done,prf:lem-pgv-substitution,prf:lem-pgv-subject-reduction-terms,prf:lem-pgv-subject-congruence,prf:thm-pgv-subject-reduction-confs}.
  \begin{lemma}%
    \label{lem:pgv-value-done}
    If $\tseq[\cs{p}]{\ty{\Gamma}}{V}{T}$, then $\cs{p}=\cs{\pbot}$, and $\minpr(\ty{\Gamma})=\minpr(\ty{T})$.
  \end{lemma}
  \input{prf/lem-pgv-value-done}

  \begin{lemma}[Substitution]%
    \label{lem:pgv-substitution}
    \hfill\\%newline before theorem statement
    If $\tseq[\cs{p}]{\ty{\Gamma},\tmty{x}{U'}}{M}{T}$ and $\tseq[\cs{q}]{\ty{\Theta}}{V}{U'}$, then $\tseq[\cs{p}]{\ty{\Gamma},\ty{\Theta}}{\subst{M}{V}{x}}{T}$.
  \end{lemma}
  \input{prf/lem-pgv-substitution}

  \begin{lemma}[Subject Reduction, $\tred$]%
    \label{lem:pgv-subject-reduction-terms}
    \hfill\\%newline before theorem statement
    If $\tseq[\cs{p}]{\ty{\Gamma}}{M}{T}$ and $\tm{M}\tred\tm{M'}$,
    then $\tseq[\cs{p}]{\ty{\Gamma}}{M'}{T}$.
  \end{lemma}
  \input{prf/lem-pgv-subject-reduction-terms}

  \begin{lemma}[Subject Congruence, $\equiv$]%
    \label{lem:pgv-subject-congruence}
    \hfill\\%newline before theorem statement
    If $\cseq[\phi]{\ty{\Gamma}}{\conf{C}}$ and $\tm{\conf{C}}\equiv\tm{\conf{C'}}$,
    then $\cseq[\phi]{\ty{\Gamma}}{\conf{C'}}$.
  \end{lemma}
  \input{prf/lem-pgv-subject-congruence}

  \begin{theorem}[Subject Reduction, $\cred$]%
    \label{thm:pgv-subject-reduction-confs}
    \hfill\\%newline before theorem statement
    If $\cseq[\phi]{\ty{\Gamma}}{\conf{C}}$ and $\tm{\conf{C}}\cred\tm{\conf{C'}}$,
    then $\cseq[\phi]{\ty{\Gamma}}{\conf{C'}}$.
  \end{theorem}
  \input{prf/thm-pgv-subject-reduction-confs}

@@

\emph{Actions} are terms which perform communication actions and which synchronise between two threads.
\begin{definition}[Actions]%
  \label{def:pgv-actions}
  A~term \emph{acts on} an endpoint $\tm{x}$ if it is $\tm{\send\;\pair{V}{x}}$, $\tm{\recv\;{x}}$, $\tm{\close\;{x}}$, or $\tm{\wait\;{x}}$. A~term is an \emph{action} if it acts on some endpoint $\tm{x}$.
\end{definition}

\emph{Ready terms} are terms which perform communication actions, either by themselves, \eg creating a new channel or thread, or with another thread, \eg sending or receiving.
It is worth mentioning that the notion of readiness presented here is akin to {live} processes introduced by Caires and Pfenning \cite{cairespfenning10,DardhaP22}, and {poised} processes introduced by Pfenning and Griffith \cite{PfenningG15} and later used by Balzer \etal \cite{balzerpfenning17,balzertoninho19}. Ready processes like live/poised processes denote processes that are ready to communicate on their providing channel.
\begin{definition}[Ready Terms]%
  \label{def:pgv-ready-actions}
  A~term $\tm{L}$ is \emph{ready} if it is of the form $\tm{\plug{E}{M}}$, where $\tm{M}$ is of the form $\tm{\new}$, $\tm{\spawn\;N}$, $\tm{\link\;\pair{x}{y}}$, or $\tm{M}$ acts on $\tm{x}$. In the latter case, we say that $\tm{L}$ is \emph{ready to act on} $\tm{x}$ or is \emph{blocked on}.
\end{definition}

@@

\begin{lemma}[Progress, $\tred$]%
  \label{lem:pgv-open-progress-terms}
  If $\tseq[\cs{p}]{\ty{\Gamma}}{M}{T}$ and $\ty{\Gamma}$ contains only session types, then:
  \begin{enumerate*}[label= (\roman*) ]
  \item $\tm{M}$ is a value;
  \item $\tm{M}\tred\tm{N}$ for some $\tm{N}$; or
  \item $\tm{M}$ is ready.
  \end{enumerate*}
\end{lemma}

@@

  \begin{definition}[Canonical Forms]%
    \label{def:pgv-canonical-forms}
    A~configuration $\tm{\conf{C}}$ is in canonical form if it is of the form
    $\tm{\res{x_1}{x'_1}{\dots\res{x_n}{x'_n}{(\child\;M_1\parallel\dots\parallel\child\;M_m\parallel\main\;N)}}}$
    where no term $\tm{M_i}$ is a value.
  \end{definition}

  \begin{lemma}[Canonical Forms]%
    \label{lem:pgv-canonical-forms}
    If $\cseq[\main]{\ty{\Gamma}}{\conf{C}}$, there exists some $\tm{\conf{D}}$ such that $\tm{\conf{C}}\equiv\tm{\conf{D}}$ and $\tm{\conf{D}}$ is in canonical form.
  \end{lemma}
  \begin{proof}
    We move any $\nu$-binders to the top using \LabTirName{SC-ResExt}, discard any superfluous occurrences of $\tm{\child\;\unit}$ using \LabTirName{SC-ParNil}, and move the main thread to the rightmost position using \LabTirName{SC-ParComm} and \LabTirName{SC-ParAssoc}.
  \end{proof}

  \begin{definition}[Normal Forms]%
    A~configuration $\tm{\conf{C}}$ is in normal form if it is of the form
    $\tm{\res{x_1}{x'_1}{\dots\res{x_n}{x'_n}{(\child\;M_1\parallel\dots\parallel\child\;M_m\parallel\main\;V)}}}$
    where each $\tm{M_i}$ is ready to act on $\tm{x_i}$.
  \end{definition}

  \begin{lemma}
    \label{lem:pgv-ready-priority}
    If $\tseq[\cs{p}]{\ty{\Gamma}}{L}{T}$ is ready to act on $\tmty{x}{S}\in\ty{\Gamma}$, then the priority bound $\cs{p}$ is some priority $\cs{o}$, \ie not $\cs{\pbot}$ or $\cs{\ptop}$.
  \end{lemma}
  \begin{proof}
    Let $\tm{L}=\tm{\plug{E}{M}}$. By induction on the structure of $\tm{E}$. $\tm{M}$ has priority $\pr({\ty{S}})$, and each constructor of the evaluation context $\tm{E}$ passes on the \emph{maximum} of the priorities of its premises. No rule introduces the priority bound $\cs{\ptop}$ on the sequent.
  \end{proof}

  \begin{theorem}[Progress, $\cred$]%
    \label{thm:pgv-closed-progress-confs}
    If $\cseq[\main]{\emptyenv}{\conf{C}}$ and $\tm{\conf{C}}$ is in canonical form, then either $\tm{\conf{C}}\cred\tm{\conf{D}}$ for some $\tm{\conf{D}}$; or $\tm{\conf{C}\equiv\conf{D}}$ for some $\tm{\conf{D}}$ in normal form.
  \end{theorem}
  \input{prf/thm-pgv-closed-progress-confs}

\begin{corollary}%
  \label{cor:pgv-closed-progress}
  If $\cseq[\phi]{\emptyenv}{\conf{C}}$, $\tm{\conf{C}}\centernot\cred$, and $\tm{\conf{C}}$ contains no endpoints, then $\tm{\conf{C}}\equiv\tm{\phi\;V}$ for some value $\tm{V}$.
\end{corollary}

>>>>

We illustrate this with two programs in PGV, \cref{ex:cycle} and \cref{ex:deadlock}. Each program contains two processes---the main process, and the child process created by $\tm{\spawn}$---which communicate using \emph{two} channels. The child process receives a unit over the channel $\tm{x}/\tm{x'}$, and then sends a unit over the channel $\tm{y}/\tm{y'}$. The main process does one of two things:
\begin{enumerate}[label= (\alph*) ]
  \item in \cref{ex:cycle}, it sends a unit over the channel $\tm{x}/\tm{x'}$, and then waits to receive a unit over the channel $\tm{y}/\tm{y'}$;
  \item in \cref{ex:deadlock}, it does these in the opposite order, which results in a deadlock.
\end{enumerate}
PGV is more expressive than GV: \cref{ex:cycle} is typeable and guaranteed to be deadlock-free in PGV, but is not typeable in GV~\cite{wadler14} and not guaranteed deadlock-free in GV's predecessor~\cite{gayvasconcelos10}. We believe PGV is a non-conservative \emph{extension} of GV, as CP can be embedded in a Kobayashi-style system~\cite{dardhaperez15extended}.

\begin{figure}
  \begin{minipage}{0.5\linewidth}
    \begin{exa}\label{ex:cycle}
      % [Cyclic Structure]%
      \[\tm{%
          \begin{array}{l}
            \letpair{x}{x'}{\new\;\unit}{}
            \\
            \letpair{y}{y'}{\new\;\unit}{}
            \\
            \spawn\left(\lambda\unit.%
            \begin{array}{l}
                {\letpair{\unit}{x'}{\recv\;{x'}}{}}
                \\
                {\letbind{y}{\send\;{\pair{\unit}{y}}}{}}
                \\
                {\andthen{\wait\;{x'}}{\close\;{y}}}
              \end{array}%
            \right);
            \\
            \underline{\letbind{x}{\send\;{\pair{\unit}{x}}}{}}
            \\
            \underline{\letpair{\unit}{y'}{\recv\;{y'}}{}}
            \\
            {\andthen{\close\;{x}}{\wait\;{y'}}}
          \end{array}
        }\]
    \end{exa}
  \end{minipage}%
  \begin{minipage}{0.5\linewidth}
    \begin{exa}\label{ex:deadlock}
      % [Deadlock]%
      \[\tm{%
          \begin{array}{l}
            \letpair{x}{x'}{\new\;\unit}{}
            \\
            \letpair{y}{y'}{\new\;\unit}{}
            \\
            \spawn\left(\lambda\unit.%
            \begin{array}{l}
                {\letpair{\unit}{x'}{\recv\;{x'}}{}}
                \\
                {\letbind{y}{\send\;{\pair{\unit}{y}}}{}}
                \\
                {\andthen{\wait\;{x'}}{\close\;{y}}}
              \end{array}%
            \right);
            \\
            \underline{\letpair{\unit}{y'}{\recv\;{y'}}{}}
            \\
            \underline{\letbind{x}{\send\;{\pair{\unit}{x}}}{}}
            \\
            {\andthen{\close\;{x}}{\wait\;{y'}}}
          \end{array}
        }\]
    \end{exa}
  \end{minipage}
\end{figure}

@@

We must show that substitution preserves priority constraints. For this, we prove~\cref{lem:pgv-value-done}, which shows that values have finished all their communication, and that any priorities in the type of the value come from the typing environment.
\begin{lem}\label{lem:pgv-value-done}
  If $\tseq[\cs{p}]{\ty{\Gamma}}{V}{T}$, then $\cs{p}=\cs{\pbot}$, and $\minpr(\ty{\Gamma})=\minpr(\ty{T})$.
\end{lem}
\input{prf/lem-pgv-value-done}

\begin{lem}\label{lem:pgv-substitution}
  %[Substitution]
  If $\tseq[\cs{p}]{\ty{\Gamma},\tmty{x}{U'}}{M}{T}$ and $\tseq[\cs{q}]{\ty{\Theta}}{V}{U'}$, then $\tseq[\cs{p}]{\ty{\Gamma},\ty{\Theta}}{\subst{M}{V}{x}}{T}$.
\end{lem}
\input{prf/lem-pgv-substitution}

\begin{lem}\label{lem:pgv-subject-reduction-terms}
  %[Subject Reduction, $\tred$]%
  If $\tseq[\cs{p}]{\ty{\Gamma}}{M}{T}$ and $\tm{M}\tred\tm{M'}$,
  then $\tseq[\cs{p}]{\ty{\Gamma}}{M'}{T}$.
\end{lem}
\input{prf/lem-pgv-subject-reduction-terms}

\begin{lem}\label{lem:pgv-subject-congruence}
  %[Subject Congruence, $\equiv$]%
  If $\cseq[\phi]{\ty{\Gamma}}{\conf{C}}$ and $\tm{\conf{C}}\equiv\tm{\conf{C'}}$,
  then $\cseq[\phi]{\ty{\Gamma}}{\conf{C'}}$.
\end{lem}
\input{prf/lem-pgv-subject-congruence}

\begin{thm}\label{thm:pgv-subject-reduction-confs}
  %[Subject Reduction, $\cred$]%
  If $\cseq[\phi]{\ty{\Gamma}}{\conf{C}}$ and $\tm{\conf{C}}\cred\tm{\conf{C'}}$,
  then $\cseq[\phi]{\ty{\Gamma}}{\conf{C'}}$.
\end{thm}
\input{prf/thm-pgv-subject-reduction-confs}

@@

\emph{Actions} are terms which perform communication actions and which synchronise between two threads.
\begin{defi}\label{def:pgv-actions}
  %[Actions]%
  A~term \emph{acts on} an endpoint $\tm{x}$ if it is $\tm{\send\;\pair{V}{x}}$, $\tm{\recv\;{x}}$, $\tm{\close\;{x}}$, or $\tm{\wait\;{x}}$. A~term is an \emph{action} if it acts on some endpoint $\tm{x}$.
\end{defi}

\emph{Ready terms} are terms which perform communication actions, either by themselves, \eg creating a new channel or thread, or with another thread, \eg sending or receiving.
It is worth mentioning that the notion of readiness presented here is akin to {live} processes introduced by Caires and Pfenning \cite{cairespfenning10,DardhaP22}, and {poised} processes introduced by Pfenning and Griffith \cite{PfenningG15} and later used by Balzer \etal \cite{balzerpfenning17,balzertoninho19}. Ready processes like live/poised processes denote processes that are ready to communicate on their providing channel.
\begin{defi}\label{def:pgv-ready-actions}
  % [Ready Terms]%
  A~term $\tm{L}$ is \emph{ready} if it is of the form $\tm{\plug{E}{M}}$, where $\tm{M}$ is of the form $\tm{\new}$, $\tm{\spawn\;N}$, $\tm{\link\;\pair{x}{y}}$, or $\tm{M}$ acts on $\tm{x}$. In the latter case, we say that $\tm{L}$ is \emph{ready to act on} $\tm{x}$ or is \emph{blocked on}.
\end{defi}

@@

\begin{lem}\label{lem:pgv-open-progress-terms}
  %[Progress, $\tred$]%
  If $\tseq[\cs{p}]{\ty{\Gamma}}{M}{T}$ and $\ty{\Gamma}$ contains only session types, then:
  \begin{enumerate*}[label= (\roman*) ]
    \item $\tm{M}$ is a value;
    \item $\tm{M}\tred\tm{N}$ for some $\tm{N}$; or
    \item $\tm{M}$ is ready.
  \end{enumerate*}
\end{lem}

@@

\begin{defi}\label{def:pgv-canonical-forms}
  %[Canonical Forms]%
  A~configuration $\tm{\conf{C}}$ is in canonical form if it is of the form
  $\tm{\res{x_1}{x'_1}{\dots\res{x_n}{x'_n}{(\child\;M_1\parallel\dots\parallel\child\;M_m\parallel\main\;N)}}}$
  where no term $\tm{M_i}$ is a value.
\end{defi}

\begin{lem}\label{lem:pgv-canonical-forms}
  % [Canonical Forms]%
  If $\cseq[\main]{\ty{\Gamma}}{\conf{C}}$, there exists some $\tm{\conf{D}}$ such that $\tm{\conf{C}}\equiv\tm{\conf{D}}$ and $\tm{\conf{D}}$ is in canonical form.
\end{lem}
\proof
We move any $\nu$-binders to the top using \LabTirName{SC-ResExt}, discard any superfluous occurrences of $\tm{\child\;\unit}$ using \LabTirName{SC-ParNil}, and move the main thread to the rightmost position using \LabTirName{SC-ParComm} and \LabTirName{SC-ParAssoc}.
\qed

\begin{defi}
  % [Normal Forms]%
  A~configuration $\tm{\conf{C}}$ is in normal form if it is of the form
  $\tm{\res{x_1}{x'_1}{\dots\res{x_n}{x'_n}{(\child\;M_1\parallel\dots\parallel\child\;M_m\parallel\main\;V)}}}$
  where each $\tm{M_i}$ is ready to act on $\tm{x_i}$.
\end{defi}

\begin{lem}\label{lem:pgv-ready-priority}
  If $\tseq[\cs{p}]{\ty{\Gamma}}{L}{T}$ is ready to act on $\tmty{x}{S}\in\ty{\Gamma}$, then the priority bound $\cs{p}$ is some priority $\cs{o}$, \ie not $\cs{\pbot}$ or $\cs{\ptop}$.
\end{lem}
\proof
Let $\tm{L}=\tm{\plug{E}{M}}$. By induction on the structure of $\tm{E}$. $\tm{M}$ has priority $\pr({\ty{S}})$, and each constructor of the evaluation context $\tm{E}$ passes on the \emph{maximum} of the priorities of its premises. No rule introduces the priority bound $\cs{\ptop}$ on the sequent.
\qed

\begin{thm}\label{thm:pgv-closed-progress-confs}
  % [Progress, $\cred$]%
  If $\cseq[\main]{\emptyenv}{\conf{C}}$ and $\tm{\conf{C}}$ is in canonical form, then either $\tm{\conf{C}}\cred\tm{\conf{D}}$ for some $\tm{\conf{D}}$; or $\tm{\conf{C}\equiv\conf{D}}$ for some $\tm{\conf{D}}$ in normal form.
\end{thm}
\input{prf/thm-pgv-closed-progress-confs}

\begin{cor}\label{cor:pgv-closed-progress}
  If $\cseq[\phi]{\emptyenv}{\conf{C}}$, $\tm{\conf{C}}\centernot\cred$, and $\tm{\conf{C}}$ contains no endpoints, then $\tm{\conf{C}}\equiv\tm{\phi\;V}$ for some value $\tm{V}$.
\end{cor}
